---
title: "Compute UVP diversity metrics"
format:
  html:
    toc: true
editor: visual
---

```{r}
#| output: false
source("utils.R")
```

## Read UVP data

```{r}
load("data/02.tara_uvp.Rdata")
```

## Clean data

### Depth

Keep only organisms above the depth at which we want to predict poc export, i.e. 1000 m.

```{r}
o <- o %>% filter(depth < 1000)
```

### Taxa

There should be nothing to do here as we use the same taxonomy as the UVP5 dataset.

```{r}
# List of taxa
taxa <- o %>% pull(taxon) %>% unique() %>% sort()
taxa
```

### Profile

Compute the number of objects per profile and keep only profiles that have more than `r n_min_uvp` objects.

```{r}
profiles %>%
  ggplot() +
  geom_polygon(data = world, aes(x = lon, y = lat, group = group), fill = "gray") +
  geom_point(aes(x = lon, y = lat, colour = n_obj > n_min_uvp)) +
  scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0)) +
  coord_quickmap()
```

```{r}
# Keep only profiles with enough objects
profiles <- profiles %>% filter(n_obj > n_min_uvp) %>% select(-n_obj)

# Drop objects that do not belong to these profiles
o <- o %>% filter(sample_ids %in% profiles$sample_ids)
```

We have `r nrow(o)` objects belonging to `r nrow(profiles)` .

### Features

Some features have a unique value for all individuals and other are missing for many individuals. Let’s remove them.

[TODO: check why `circ.` is binary.]{style="color: red;"}

```{r}
# Select features
x <- o %>% select(area:areai)

# Remove variables with zero variance
feats <- x %>%
  summarise_all(var, na.rm = TRUE) %>%
  pivot_longer(cols = everything()) %>%
  filter(value > 0) %>%
  pull(name)

# Exclude some features with many missing values
feats <- feats[!str_detect(feats, "exc")]

# Exclude circ. as it is binary
# TODO check why
feats <- feats[feats != "circ."]

x <- x %>% select(all_of(feats))
```

Plot features distributions.

```{r}
#| column: screen
#| out-width: 60%
#| fig-align: center
#| fig-width: 10
#| fig-height: 8
x %>%
  pivot_longer(cols = everything()) %>%
  ggplot() +
  geom_histogram(aes(x = value), bins = 50) +
  facet_wrap(~name, scales = "free")
```

For a PCA, features should be normaly-distributed. Let’s apply some transformation to get closer to normal distribution:

-   mask extreme values

-   normalize using the Yeo-Johnson transformation

-   replace missing values by the mean of each column

```{r}
x_norm <- x %>%
  # remove the most extreme high values
  mutate_all(mask_extreme, percent = c(0, 0.5)) %>%
  # normalise using the Yeo-Johnson transformation
  mutate_all(yeo_johnson) %>%
  mutate_all(as.numeric)

# Replace NA by average of each column
for (col in names(x_norm)) {
  x_norm[[col]][is.na(x_norm[[col]])] <- mean(x_norm[[col]], na.rm=TRUE)
}
```

Plot normalized features.

```{r}
#| column: screen
#| out-width: 60%
#| fig-align: center
#| fig-width: 10
#| fig-height: 8
x_norm %>%
  pivot_longer(cols = everything()) %>%
  ggplot() +
  geom_histogram(aes(x = value), bins = 50) +
  facet_wrap(~name, scales = "free")
```

## Taxonomic diversity

Compute taxonomic diversity indices:

-   Taxonomic richness

-   Shannon Diversity

-   Pielou Evenness

```{r}
# Generate a contingency table as a matrix to feed to vegan
cont <- o %>%
  count(sample_ids, taxon) %>%
  pivot_wider(names_from = "taxon", values_from = "n", values_fill = 0) %>%
  as.data.frame() %>%
  column_to_rownames(var = "sample_ids") %>%
  as.matrix()

# Compute diversity metrics
ta_div_prof <- tibble(
  sample_ids = rownames(cont),
  ta_ric = specnumber(cont),
  ta_shannon = diversity(cont, index = "shannon"),
  ta_pielou = ta_shannon/log(ta_ric)
) %>%
  left_join(profiles, by = join_by(sample_ids)) %>%
  select(sample_ids, lon, lat, everything())


# Store results with table of profiles
profiles <- profiles %>% left_join(ta_div_prof, by = join_by(sample_ids, lon, lat))
```

Plot taxonomic diversity metrics.

```{r}
#| column: screen
#| out-width: 60%
#| fig-align: center
ggmap(
  profiles, 
  "ta_ric", 
  type = "point"
  )

ggmap(
  profiles, 
  "ta_shannon", 
  type = "point"
  )

ggmap(
  profiles, 
  "ta_pielou", 
  type = "point"
  )
```

## Trophic diversity

## Morphological diversity

### Morphospace

#### Build

Let’s feed the features to a PCA to build a morphospace.

```{r}
# We need to use "scale.unit = TRUE" to center-scale all feature
m_space <- FactoMineR::PCA(x_norm, scale.unit = TRUE, graph=FALSE)
```

#### Eigenvalues

Plot the eigenvalues.

```{r}
eig <- m_space$eig %>%
  as.data.frame() %>%
  rownames_to_column(var = "comp") %>%
  as_tibble() %>%
  mutate(
    comp = str_remove(comp, "comp "),
    comp = as.numeric(comp),
    comp = as.factor(comp)
    ) %>% 
  rename(var = `percentage of variance`, cum_var = `cumulative percentage of variance`)

eig %>%
  ggplot() +
  geom_col(aes(x = comp, y = eigenvalue)) +
  theme_classic() +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "PC", y = "Eigenvalue")
```

Most of the variance is captured by the first two axes (`r format(eig$var[1]/100, digits = 2)` and `r format(eig$var[2]/100, digits = 2)` respectively).

[TODO: PC selection.]{style="color: red;"}

#### Features and axis definition

Let’s now plot the first two axes.

```{r}
#| column: screen
#| out-width: 60%
#| fig-align: center
#| fig-width: 8
#| fig-height: 8
plot(m_space, choix="var", axes = c(1, 2))
```

```{r}
#| column: screen
#| out-width: 60%
#| fig-align: center
#| fig-width: 8
#| fig-height: 8
plot(m_space, choix="var", axes = c(2, 3))
```

-   PC1: big objects in positive values, small objects in negative values.

-   PC2: clear (i.e. transparent) objects in positive values, dark (i.e. opaque) objects in negative values.

-   PC3: harder to tell, something with gray levels.

#### Individuals

Let’s extract the coordinates of individuals in the morphospace.

```{r}
## Get coordinates of individuals
inds <- m_space$ind$coord %>% as_tibble()
# Set nice names for columns
colnames(inds) <- str_c("dim", paste(c(1:ncol(inds))))
# And join with initial dataframe of objects
o <- o %>%
  select(sample_ids:lat) %>%
  bind_cols(inds)
```

We can not plot the position of objects in the morphospace, coloured per profile.

```{r}
## Plot invidivuals with profile as colour
o %>%
  ggplot() +
  geom_point(aes(x = dim1, y = dim2, colour = sample_ids), show.legend = FALSE, size = 0.5, alpha = 0.5) 
```

### Diversity

#### Morphospace features

We can collect the position of objects in the morphospace to summarise the morphological diversity of each profile.

```{r}
# Compute mean and variance of dim1 and dim2 per profile
m_div_prof <- o %>% 
  group_by(sample_ids, lon, lat) %>% 
  summarise(across(dim1:dim2, list(mean = mean, var = var))) %>% 
  ungroup()

# And store this with profiles data
profiles <- profiles %>% left_join(m_div_prof, by = join_by(sample_ids, lon, lat))
```

And we can plot maps of mean `dim1` and `dim2` values for each profile.

```{r}
#| column: screen
#| out-width: 60%
#| fig-align: center
ggmap(
  profiles, 
  "dim1_mean", 
  type = "point", 
  palette = div_pal
  )
```

```{r}
#| column: screen
#| out-width: 60%
#| fig-align: center
ggmap(
  profiles, 
  "dim2_mean", 
  type = "point", 
  palette = div_pal
  )
```

We can also look at variance within profiles.

```{r}
#| column: screen
#| out-width: 60%
#| fig-align: center
ggmap(
  profiles, 
  "dim1_var", 
  type = "point"
  )
```

```{r}
#| column: screen
#| out-width: 60%
#| fig-align: center
ggmap(
  profiles, 
  "dim2_var", 
  type = "point"
  )
```

#### Metrics

Multivariate morphological diversity metrics have been defined in [Beck et al. 2023](https://royalsocietypublishing.org/doi/full/10.1098/rspb.2023.2109) following the definition of multivariate functional diversity metrics in [Villeger et al. 2008](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1890/07-1206.1):

-   morphological richness

-   morphological evenness

-   morphological divergence

Computing these metrics require defining "morphs" (i.e. morphologically similar organisms) in the morphospace, i.e. using kmeans. These morphs are then used instead of species to compute morphological diversity metrics.

## Save

```{r}
save(profiles, file = "data/03.uvp_profiles.Rdata")
```
