---
title: "Predict plankton values from env."
subtitle: "Train a MBRT model to predict a few plankton variables (the best POC predictors) from env data."
author: "Thelma Panaïotis"
format:
  html:
    toc: true
    embed-resources: true
editor: visual
execute:
  cache: true
  warning: false
---

## Set-up

### Python

```{python py_set_up}
#| include: false
#| cache: false
import os
import pandas as pd
import numpy as np
import multiprocess as mp
import pyreadr

import matplotlib.pyplot as plt
import cartopy.crs as ccrs

from mbtr.mbtr import MBT
from sklearn import metrics
from sklearn.model_selection import train_test_split, KFold, ParameterGrid

random_state = 12
```

### R

```{r r_set_up}
#| include: false
#| warning: false
#| cache: false
source("utils.R")
output_filename <- "data/10.all_preds_plankton_best_from_env.Rdata"
```

::: callout-caution
-   Consider transforming input data for better results?
:::

## Read data

Define variables:

-   metadata

-   response variables: plankton variables that are the best POC predictors

-   explanatory variables: env variables as in Wang et al., 2023

```{r read_data}
load("data/04.all_data.Rdata")

## Types of variables
# Metadata
meta_vars <- df %>% select(lon:datetime) %>% colnames()
# Response variables, here we use only best POC predictors
#resp_vars <- df %>% select(ta_ric:m_eve) %>% colnames()
resp_vars <- c("mo_ric", "ta_div", "ta_eve")
# Explanatory variables
#exp_vars <- df %>% select(temperature_mean:par_sd) %>% colnames()
exp_vars <- df %>% 
  select(
    temperature_mean,
    silicate_mean,
    z_eu_mean,
    oxygen_mean
    ) %>% 
  colnames()

# subsample
#df <- df %>% slice_sample(prop = 0.2)
```

Inspect response data.

```{r plot_resp}
#| fig-column: screen-inset
#| out-width: 100%
#| fig-width: 15
#| fig-height: 4
df %>% 
  select(all_of(resp_vars)) %>% 
  pivot_longer(everything()) %>% 
  ggplot() +
  geom_histogram(aes(x = value)) +
  facet_wrap(~name, scales = "free")
```

## Split data

Split data into training VS testing

```{r split_train_test}
# Train VS test
df_split <- initial_split(df, prop = 0.8)
df_train <- training(df_split)
df_test <- testing(df_split)
```

Split data into response VS predictor

```{python split_exp_resp}
# Explanatory variables
X_train = r.df_train[r.exp_vars].to_numpy()
X_test = r.df_test[r.exp_vars].to_numpy()

# Response variables
Y_train = r.df_train[r.resp_vars].to_numpy()
Y_test = r.df_test[r.resp_vars].to_numpy()
```

## MBTR

### Gridsearch

```{python gridsearch}
# define the grid of parameters to search over
grid = {
    'n_boosts' : [100], 
    'lambda_weights' : [0.001],
    'lambda_leaves' : [0.001],
    'early_stopping_rounds' : [10],
    'learning_rate' : [0.1, 0.05, 0.025],
    'min_leaf' : [1, 4, 8]
}


# initialise a data splitter into 5 folds
n_splits = 3
cv = KFold(n_splits = n_splits, shuffle = True, random_state = random_state)

res = []

def fit_cv(params):
  
    # initialise
    val_loss = np.zeros((params['n_boosts'],n_splits))
    k = 0

    for train,val in cv.split(X_train, Y_train):
        # fit model
        MBT_model = MBT(**params)
        MBT_model.fit(X_train[train], Y_train[train])
        # compute validation curve
        for i in range(params['n_boosts']):
            Y_pred = MBT_model.predict(X_train[val], n=i+1)
            val_loss[i,k] = metrics.mean_squared_error(
                Y_train[val],
                Y_pred
            )
        k += 1

    # store result
    res.append({**params, 'val_loss': val_loss})
    
    return(res)
  
## Run this in parallel
# number of cores is minimum between:
# - nb of available cores - 2
# - nb of grids to test
n_cores = min(mp.cpu_count() - 2, len(list(ParameterGrid(grid)))) 
with mp.Pool(n_cores) as pool:
    res = pool.map(fit_cv, ParameterGrid(grid))

# Unlist top level
res = [l for li in res for l in li]
save_res = res

```

Plot results

```{python plot_grid_res}
#| warning: false
n_trees = np.arange(1, grid['n_boosts'][0]+1)

fig,axs = plt.subplots(1, len(grid['min_leaf']))
s = 0
for min_leaf in grid['min_leaf']:
    res_subset = [r for r in res if (r['min_leaf'] == min_leaf)]
    
    for i in range(len(res_subset)):
        axs[s].plot(n_trees, np.mean(res_subset[i]['val_loss'], axis=1),
            label=f"lr: {res_subset[i]['learning_rate']:.3f}")
        axs[s].legend()
        axs[s].set_title(f"min_leaf: {min_leaf}")
        #axs[s].set_ylim([0.001, 0.0035])
    s += 1

plt.show()
```

### Get best params with R

```{r get_best_params}
res <- py$res
# Convert results to dataframe
df_res <- tibble()
for (i in 1:length(res)){
  df_res <- rbind(df_res, data.frame(res[i]))
}

# Compute mean of val_loss across CV folds
df_res <- df_res %>% 
  mutate(val_loss = select(., val_loss.1:val_loss.3) %>% rowMeans()) %>% 
  select(-c(val_loss.1:val_loss.3))

# Get lower val_loss
best_params <- df_res %>% 
  arrange(val_loss) %>% 
  select(-val_loss) %>% 
  dplyr::slice(1) %>% 
  pivot_longer(cols = everything())
best_params
```

### Refit the model

```{python refit}
# convert dataframe to dict
best_params = r.best_params
best_params = best_params.set_index('name')
best_params = best_params['value'].to_dict()

# Set n_boosts as an integer
best_params['n_boosts'] = int(best_params['n_boosts'])
# Disable early stopping
best_params['early_stopping_rounds'] = 100

# Refit model
MBT_model = MBT(**best_params)
MBT_model.fit(X_train, Y_train)

# Predict test set to get metrics
Y_test_pred = MBT_model.predict(X_test, n=best_params['n_boosts'])
print(f"Multiple R2 on test set : {metrics.r2_score(Y_test, Y_test_pred):.2f}")
```

```{r plot_pred_vs_truth}
#| out-width: 100%
#| fig-width: 16
#| fig-height: 4
# Get predictions and true values, as vertical DF to bind together
Y_test_pred <- py$Y_test_pred %>% 
  as_tibble(.name_repair = "minimal") %>% 
  set_names(resp_vars) %>% 
  pivot_longer(all_of(resp_vars), names_to = "variable", values_to = "pred")

Y_test <- py$Y_test %>% 
  as_tibble(.name_repair = "minimal") %>% 
  set_names(resp_vars) %>% 
  pivot_longer(all_of(resp_vars), names_to = "variable", values_to = "truth")

Y_test %>% 
  bind_cols(Y_test_pred %>% select(-variable)) %>% 
  ggplot() +
  geom_point(aes(x = truth, y = pred)) +
  geom_abline(slope = 1, intercept = 0, colour = "red") +
  facet_wrap(~variable, scales = "free")
```

### Predict all data

Predict all data to get all residuals.

```{python pred_all}
X_all = np.concatenate((X_train, X_test))
Y_all = np.concatenate((Y_train, Y_test))
Y_all_pred = MBT_model.predict(X_all, n=best_params['n_boosts'])
```

### Compute residuals and save them

```{r get_resids_and_save}
#| fig-column: screen-inset
#| out-width: 100%
#| fig-width: 10
#| fig-height: 8

# Get predictions for all dataset
all_preds <- py$Y_all_pred %>% 
  as_tibble(.name_repair = "minimal") %>% 
  set_names(paste0("pred_", resp_vars))

# Generate a df of residuals
all_resids <- py$Y_all - py$Y_all_pred %>% 
  as_tibble(.name_repair = "minimal") %>% 
  set_names(paste0("resid_", resp_vars))

# Add metadata
plankton_res <- df %>% 
  select(all_of(meta_vars), all_of(resp_vars)) %>% 
  bind_cols(all_preds) %>% 
  bind_cols(all_resids)

# Save
save(plankton_res, file = output_filename)
```
