---
title: "Predicting POC export from plankton diversity"
subtitle: "Calipsocean meeting"
author: "Thelma Panaïotis"
date: "05/03/2024"
date-format: "DD/MM/YYYY"
format: 
  revealjs:
    theme: simple
    #footer: ""
    slide-number: c
editor: visual
---

```{r}
#| output: false
#| echo: false
source("utils.R")

load("data/00.carbon_data.Rdata")
load("data/01.all_env.Rdata")
load("data/03.uvp_profiles.Rdata")
load("data/04.all_data.Rdata")
rm(coast)
```

# Data

## POC

POC export at 1000 m, from [Wang et al. 2023](https://www.nature.com/articles/s41586-023-06772-4).

```{r}
ggmap(df = df_c, var = "poc", type = "raster")
```

## Environment

When you click the **Render** button a document will be generated that includes:

-   Content authored with markdown
-   Output from executable code

## Plankton

UVP5 dataset

```{r}
ggplot(profiles, aes(x = lon, y = lat)) +
  geom_polygon(data = world, aes(group = group), fill = "gray") +
  geom_point(size = 0.5) +
  coord_quickmap(expand = FALSE)
```

## Plankton - Taxonomic Diversity

::: columns
::: {.column width="30%"}
-   species richness

-   Shannon index

-   Pielou’s evenness
:::

::: {.column width="70%"}

```{r}
ggmap(profiles, var = "ta_shannon", type = "point") + labs(colour = "Shannon")
```
:::
:::

## Plankton - Trophic Diversity

::: columns
::: {.column width="40%"}
Left column
:::

::: {.column width="60%"}
Right column
:::
:::

## Dataset

`r nrow(df)` data points

```{r}
ggmap(df, var = "poc", type = "point")
```

# Model

## Code

When you click the **Render** button a presentation will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```
